Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-17 paper.qmd
Restructured video links table

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and (hyper-)~parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

<!-- - Connection between NLDR and MDS-->
Historically, low-dimensional (\kD{}) representations of high-dimensional (\pD{}) data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.)
<!-- Does JASA allow footnotes? If so this is a footnote, otherwise the punctuation has gone a bit rogue - can't have an entire sentence in brackets surely --> The \kD{} representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{} layout is constructed by minimizing a stress function that differences distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

<!-- Non-expert comment: This to me is a leap from MDS to NLDR. Is NLDR an MDS? I don't think so, but I'm missing the link -->
<!-- Suggested rewording: "To better capture specific structures potentially existing in \pD{}, many new methods for NLDR have emerged in recent years." -->
Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. Both tNSE and UMAP can be considered to produce the \kD{} representation by minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)~parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

<!-- I'm not thrilled about H_h but I guess it can stay because changing it everywhere is a pain -->

@@

<!-- In Figure 4, is the "sparse space" in the middle of each cluster (i.e. the holes) or is it the gap between clusters? If it's between clusters, I'd argue tSNE is making this gap smaller than in reality, which I don't think the caption currently conveys. -->

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of nonlinear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to \\fD{} (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)~parameter choice."
#| fig-width: 8
#| fig-height: 2

@@

@fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default (hyper-)~parameters for each method are used. In each layout we can see an accurate representation where all five clusters are visible, although with varying degrees of separation.

@@

@fig-one-dens_clust-error (a2, b2) shows a \gD{} layout for tSNE created using the default (hyper-)~parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their \\fD{} model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end."

@@

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different (hyper-)~parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

@fig-pbmc-rmse shows RMSE across a range of binwidths ($a_1$) for each of the layouts in @fig-NLDR-variety. The layouts were made using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with various (hyper-)~parameter settings. Lines are color coded to match the color of the layouts shown on the right. Lower RMSE indicates the better fit. Using a range of binwidths shows how the model changes, with possibly the best model being one that is universally low RMSE across all binwidths. It can be seen that layout f is sub-optimal with universally higher RMSE. Layout a, the published one, is better but it is not as good as layouts b, d, or e. With some imagination layout d perhaps shows three barely distinguishable clusters. Layout e shows three, possibly four, clusters that are more separated. The choice reduces from eight to these two. Layout d has slightly better RMSE when the $a_1$ is small, but layout e beats it at larger values. Thus we could argue that layout e is the most accurate representation of the cluster structure, of these eight.

@@

@fig-mnist-rmse compares the fit of six layouts computed using UMAP (b), PHATE (c), TriMAP (d), PaCMAP (e) with default (hyper-)~parameter setting and two tSNE runs, one with default (hyper-)~parameter setting (a) and the other changing perplexity to $89$ (f). The layouts are reasonably similar in that they all have the observations in a single blob. Some (b, c) have a more curved shape than others. Layout e is the most different having a linear shape, and a single very large outlier. Both a and f have a small clump of points perhaps slightly disconnected from the other points, in the lower to middle right.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)~parameter choice(s). It depends on conceptualizing the \gD{} layout as a model, allowing for the creation of a wireframe representation of the model that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

@@

```{r}
#| label: tbl-links
#| tbl-cap: "Videos of the langevitour animations and the linked plots."
# Video links in the paper

@@

# Create the table
kable(links_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:3, width = c("1.5cm", "4.5cm", "9cm"))
```

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and hyper-parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

<!-- - Connection between NLDR and MDS-->
Historically, low-dimensional (\kD{}) representations of high-dimensional (\pD{}) data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.) The \kD{} representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{} layout is constructed by minimizing a stress function that differences distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. Both tNSE and UMAP can be considered to produce the \kD{} representation by minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with hyper-parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

@@

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the \\gD{} fitted model into \\pD{}. Two projections of the \\pD{} fitted model overlaying the data are shown in b, c. The fit is reasonbly tight with the data in one cluster (top one in b), but slightly less so in the other cluster probably because it is in $3-\\text{D}$. Notice also that, in the \\gD{} layout the two clusters have internal gaps which creates a model with some holes. This lacy pattern happens regardless of the hyper-parameter choice, but this doen't severely impact the \\pD{} model representation."
#| fig-width: 8
#| fig-height: 2

@@

@fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters for each method are used. In each layout we can see an accurate representation where all five clusters are visible, although with varying degrees of separation.

@@

@fig-one-dens_clust-error (a2, b2) shows a \gD{} layout for tSNE created using the default hyper-parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet.

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Exploring the effect of density on the NLDR layout using a C-shaped structure with different density in each end. The sparse end of the structure shows a high model error, while the dense end shows a low model error. The top part of the triangular shaped with linked brushing. The tSNE layout, with dense points (a1-a3) are colored according to their \\fD{} model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end."

@@

In this data there are $2622$ single cells and $1000$ gene expressions (variables). Following their pre-processing, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods, and different hyper-parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout published in @chen2024. This layout suggests that the data has three very well separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data.

@fig-pbmc-rmse shows RMSE across a range of binwidths ($a_1$) for each of the layouts in @fig-NLDR-variety. The layouts were made using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with various hyper-parameter settings. Lines are color coded to match the color of the layouts shown on the right. Lower RMSE indicates the better fit. Using a range of binwidths shows how the model changes, with possibly the best model being one that is universally low RMSE across all binwidths. It can be seen that layout f is sub-optimal with universally higher RMSE. Layout a, the published one, is better but it is not as good as layouts b, d, or e. With some imagination layout d perhaps shows three barely distinguishable clusters. Layout e shows three, possibly four, clusters that are more separated. The choice reduces from eight to these two. Layout d has slightly better RMSE when the $a_1$ is small, but layout e beats it at larger values. Thus we could argue that layout e is the most accurate representation of the cluster structure, of these eight.

@@

@fig-mnist-rmse compares the fit of six layouts computed using UMAP (b), PHATE (c), TriMAP (d), PaCMAP (e) with default hyper-parameter setting and two tSNE runs, one with default hyper-parameter setting (a) and the other changing perplexity to $89$ (f). The layouts are reasonably similar in that they all have the observations in a single blob. Some (b, c) have a more curved shape than others. Layout e is the most different having a linear shape, and a single very large outlier. Both a and f have a small clump of points perhaps slightly disconnected from the other points, in the lower to middle right.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and hyper-parameter choice(s). It depends on conceptualizing the \gD{} layout as a model, allowing for the creation of a wireframe representation of the model that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

@@

```{r}
#| label: tbl-links
#| tbl-pos: H
#| tbl-cap: "Videos of the langevitour animations and the linked plots."
# Video links in the paper

@@

# Create the table
kable(links_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("1.5cm", "4.5cm"))
```
